---
title: "Data Basics"
author: Zach del Rosario
date: 2020-05-03
output: github_document
time: 10
reading: 0
editor_options: 
  markdown: 
    wrap: 72
---

# Data: Basics

*Purpose*: When first studying a new dataset, there are very simple
checks we should perform first. These are those checks.

Additionally, we'll have our first look at the *pipe operator*, which
will be super useful for writing code that's readable.

*Reading*: (None)

```{r setup}
library(tidyverse)
```

```{r options, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

## First Checks

<!-- -------------------------------------------------- -->

### **q0** Run the following chunk:

*Hint*: You can do this either by clicking the green arrow at the
top-right of the chunk, or by using the keybaord shortcut `Shift` +
`Cmd/Ctrl` + `Enter`.

```{r }
head(iris)
```

This is a *dataset*; the fundamental object we'll study throughout this
course. Some nomenclature:

-   The `1, 2, 3, ...` on the left enumerate the **rows** of the dataset
-   The names `Sepal.Length`, `Sepal.Width`, `...` name the **columns**
    of the dataset
-   The column `Sepal.Length` takes **numeric** values
-   The column `Species` takes **string** values

### **q1** Load the `tidyverse` and inspect the `diamonds` dataset. What do the

`cut`, `color`, and `clarity` variables mean?

*Hint*: You can run `?diamonds` to get information on a built-in
dataset.

```{r load-packages, eval=FALSE}
?diamonds
```

    cut: quality of the cut
    color: diamond color, from D to J (best to worst)
    clarity: measures how clear the diamond is

### **q2** Run `glimpse(diamonds)`; what variables does `diamonds` have?

```{r}
glimpse(diamonds)
```

    The variables of `diamond` are: carat, cut, color, clarity, depth, table, price, x, y, z

### **q3** Run `summary(diamonds)`; what are the common values for each of the variables? How widely do each of the variables vary?

*Hint*: The `Median` and `Mean` are common values, while `Min` and `Max`
give us a sense of variation.

```{r q3-task}
summary(diamonds)
```

**Observations**:

-   All the summaries come with a price associated with each value. For
    some of the discrete data, the categories are listed and the mean
    price is listed. For continuous values, a statistical breakdown is
    provided with the min/max and mean and other values.

-   Carat: the carat varies from 0.2 to 5.01, and the prices of this
    range are from 326 to 18823.

-   Cut: there are 5 values representing the quality of the cut, and as
    the cut gets bigger, the price increases (range of 1610 to 21551)

-   Color: Some colors are more valuable than others, but there is no
    sorted order other than the price. The price ranges from 2808 to
    11292

-   Clarity: The clearer the stone, the higher the worth. The price
    ranges from 2531 to 13065

-   Depth: The depth is a continuous value. The price based on the range
    of depths ranges from 43 to 79.

-   Table: what the hell is this

-   the x value ranges from 0 to 10.74

-   y ranges from 0 to 58.9

-   z ranges from 0 to 31.8

You should always analyze your dataset in the simplest way possible,
build hypotheses, and devise more specific analyses to probe those
hypotheses. The `glimpse()` and `summary()` functions are two of the
simplest tools we have.

## The Pipe Operator

<!-- -------------------------------------------------- -->

Throughout this class we're going to make heavy use of the *pipe
operator* `%>%`. This handy little function will help us make our code
more readable. Whenever you see `%>%`, you can translate that into the
word "then". For instance

```{r pipe-example}
diamonds %>%
  group_by(cut) %>%
  summarize(carat_mean = mean(carat))
```

Would translate into the tiny "story"

-   Take the `diamonds` dataset, *then*
-   Group it by the variable `cut`, *then*
-   summarize it by computing the `mean` of `carat`

*What the pipe actually does*. The pipe operator `LHS %>% RHS` takes its
left-hand side (LHS) and inserts it as an the first argument to the
function on its right-hand side (RHS). So the pipe will let us take
`glimpse(diamonds)` and turn it into `diamonds %>% glimpse()`.

### **q4** Use the pipe operator to re-write `summary(diamonds)`.

```{r q4-task}
diamonds %>% summary()
```

## Reading Data

<!-- -------------------------------------------------- -->

So far we've only been looking at built-in datasets. Ultimately, we'll
want to read in our own data. We'll get to the art of loading and
*wrangling* data later, but for now, know that the `readr` package
provides us tools to read data. Let's quickly practice loading data
below.

### **q5** Use the function `read_csv()` to load the file `"./data/tiny.csv"`.

```{r q5-task}

df_q5 <-
  read_csv("./data/tiny.csv")
```

<!-- include-exit-ticket -->

# Exit Ticket

<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit
ticket survey**, [linked
here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-data00-basics-assignment.Rmd).
