---
title: "Vis: Histograms"
author: Zach del Rosario
date: 2020-05-22
output: github_document
time: 30
reading: 20
---

# Vis: Histograms

*Purpose*: *Histograms* are a key tool for EDA. In this exercise we'll get a little more practice constructing and interpreting histograms and densities.

*Reading*: [Histograms](https://rstudio.cloud/learn/primers/3.3) *Topics*: (All topics) *Reading Time*: \~20 minutes

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

### **q1** Using the graphs generated in the chunks `q1-vis1` and `q1-vis2` below, answer:

-   Which `class` has the most vehicles?
    -   SUV's
-   Which `class` has the broadest distribution of `cty` values?
    -   subcompact - flattest curve
-   Which graph---`vis1` or `vis2`---best helps you answer each question?
    -   vis1 is good for answering question 1, vis2 is good for question 2

```{r q1-vis1}
## NOTE: No need to modify
mpg %>%
  ggplot(aes(cty, color = class)) +
  geom_freqpoly(bins = 10)
```

```{r q1-vis2}
## NOTE: No need to modify
mpg %>%
  ggplot(aes(cty, color = class)) +
  geom_density()
```

In the previous exercise, we learned how to *facet* a graph. Let's use that part of the grammar of graphics to clean up the graph above.

### **q2** Modify `q1-vis2` to use a `facet_wrap()` on the `class`. "Free" the vertical axis with the `scales` keyword to allow for a different y scale in each facet.

```{r q2-task}
mpg %>%
  ggplot(aes(cty, color = class)) +
  geom_density() +
  facet_wrap(~class, scales = "free_y")
```

In the reading, we learned that the "most important thing" to keep in mind with `geom_histogram()` and `geom_freqpoly()` is to *explore different binwidths*. We'll explore this idea in the next question.

### **q3** Analyze the following graph; make sure to test different binwidths. What patterns do you see? Which patterns remain as you change the binwidth?

```{r q3-task}
## TODO: Run this chunk; play with differnet bin widths
diamonds %>%
  filter(carat < 1.1) %>%

  ggplot(aes(carat)) +
  geom_histogram(binwidth = 0.1, boundary = 0.005) +
  scale_x_continuous(
    breaks = seq(0, 1, by = 0.1)

  )
```

**Observations** - Write your observations here!

-   with the smaller bin widths (around 0.01), the pattern clearly shows that there is a huge spike in number of diamonds with more "whole decimal" carats, like 0.3, 0.5, 0.7, etc. There are very little diamonds of carat sizes slightly less than these whole decimals, but a lot of carat sizes slightly more than them. The amount decreases exponentially.

-   With a bin width of 0.05, the exponential decreases aren't noticable, but the peaks are generally noticable

-   with any larger of a bin width, you cannot tell where the peaks are

<!-- include-exit-ticket -->

# Exit Ticket

<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis02-histograms-assignment.Rmd).
