---
title: "Vis: Boxplots and Counts"
author: Zach del Rosario
date: 2020-06-04
output: github_document
dep: e-data02-derive
time: 20
reading: 20
editor_options: 
  markdown: 
    wrap: 72
---

# Vis: Boxplots and Counts

*Purpose*: *Boxplots* are a key tool for EDA. Like histograms, boxplots
give us a sense of "shape" for a distribution. However, a boxplot is a
*careful summary* of shape. This helps us pick out key features of a
distribution, and enables easier comparison of different distributions.

*Reading*: [Boxplots and
Counts](https://rstudio.cloud/learn/primers/3.4) *Topics*: (All topics)
*Reading Time*: \~20 minutes

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

A *subtle point* from the primer is that we can use `dplyr` to generate
new facts about our data, then use `ggplot2` to visualize those facts.
We'll practice this idea in a number of ways.

### **q1** Use a `cut_*` verb to create a categorical variable out of `carat`. Tweak the settings in your cut and document your observations.

*Hint*: Recall that we learned about
`cut_interval, cut_number, cut_width`. Take your pick!

```{r q1-task}
diamonds %>%
  mutate(carat_cut = cut_interval(x = carat, n = 10)) %>%
  ggplot(aes(x = carat_cut, y = price)) +
  geom_boxplot()
```

**Observations** - Write your observations here! - Make sure to
distinguish between the trend in median (bar), quantiles (box edges),
and whiskers (vertical lines)

-   From 0.2 to about 3 carats, the median price tends to increase in a
    positive trend

-   The price distribution of diamonds from 0.2 to about 1.64 carats
    tends to be skewed right, since there are more outliers above the
    box. From about 1.64 to 3 carats, the distribution tends to be
    skewed left since there are more outliers and the whisker is longer
    below the box. Beyond 3 carats, the distribution is more normal, but
    skewed right if skewed at all. This was all determined through
    looking mainly at the length of the whiskers above/below the box,
    and a little bit through looking at the position of the median
    within the IQR.

### **q2** The following code visualizes the count of diamonds of *all* carats according to their cut and color. Modify the code to consider *only* diamonds with `carat >= 2`. Does the most common group of cut and color change?

```{r q2-ref}
## NOTE: No need to modify; run and inspect
diamonds %>%
  count(cut, color) %>%
  ggplot(aes(cut, color, fill = n)) +
  geom_tile()
```

Modify the following code:

```{r q2-task}
## TODO: Modify the code below to consider only diamonds with carat >= 2
diamonds %>%
  filter(carat >= 2) %>%
  count(cut, color) %>%
  ggplot(aes(cut, color, fill = n)) +
  geom_tile()
```

**Observations**: - Did the most common group of cut and color change?

-   yes, across all diamonds, ideal diamonds with color near G on the
    spectrum tend to be the most common.

-   if we ignore diamonds than 2 carats, then Premium diamonds with
    colors closer to I on the spectrum tend to be more common.

-   This tells me that smaller diamonds tend to be polished to ideal -
    or possibly that polishing to ideal causes a smaller diamond since
    you're grinding so much off

### **q3** The following plot has overlapping x-axis labels. Use a verb from the reading to `flip` the coordinates and improve readability.

```{r q3-task}
mpg %>%
  ggplot(aes(manufacturer, hwy)) +
  geom_boxplot() +
  coord_flip()
```

This is a simple---but important---trick to remember when visualizing
data with many categories.

<!-- include-exit-ticket -->

# Exit Ticket

<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit
ticket survey**, [linked
here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis03-boxplots-assignment.Rmd).
