---
title: "RMS Titanic"
author: "Lily Jiang"
date: 2023-01-30
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|-------------|------------------------------------------------------------------------------------------------------------------|----------------------------------------------------------------------------------------------------------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}

glimpse(df_titanic)
```

**Observations**:

-   Class, Sex, Age, Survived, and number of each (n)

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

-   Write your observations here
-   Are there any differences?
    -   Wikipedia says 2224 people were on the boat, while this dataset says there are only 2201
-   If yes, what might account for those differences?
    -   According to Wikipedia, there was confusion with the passenger list because some people canceled last minute and some people were double-counted for travelling under aliases. Therefore, there isn't a universally-accurate count of passengers on the Titanic. This dataset probably is derived from a different source than the Wikipedia article, which would explain these differences.

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`

df_titanic %>%
  filter(Survived == "Yes") %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = n, fill = Sex), width = 0.5, position = "dodge")
```

**Observations**:

-   In the non-crew passengers, overall more female passengers survived. This is likely due to the general rule of "women and children first".
-   In 3rd class, about the same number of male and female passengers survived
-   Amongst the crew, far more males survived. This could likely be due to the fact that the crew was composed mostly of males due to the gender norms of the time and overall hard labor required to maintain a ship.

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot() +
  geom_col(aes(x = Class, y = Prop, fill = Sex), width = 0.5, position = "dodge")
```

**Observations**:

-   Write your observations here.
    -   All of the people in 1st and 2nd class survived, while very few people in 3rd class and of the crew survived. Class injustices at play

    -   A greater proportion of women than men survived in 3rd class and the crew.
-   Is there anything *fishy* going on in your plot?
    -   The 100% survival rate in 1st and 2nd classes, whereas the 3rd class had a very small proportion of survivors

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot() +
  geom_col(aes(x = Class, y = Prop, fill = Sex), width = 0.5, position = "dodge") +
  facet_grid(~ Age)
```

**Observations**:

-   All of the children in 1st and 2nd class survived. Most women in 1st and 2nd class survived
-   A very few proportion of men survived across all classes
-   If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
    -   100% of the children in 1st and 2nd class survived, while only some of the adults (largely women) survived. This explains that not *all* of the 1st and 2nd class members survived.. the high proportion comes from all the children.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
