---
title: "Antibiotics"
author: "(Lily Jiang)"
date: 2022-3-5
output:
  github_document
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: Creating effective data visualizations is an *iterative*
process; very rarely will the first graph you make be the most
effective. The most effective thing you can do to be successful in this
iterative process is to *try multiple graphs* of the same data.

Furthermore, judging the effectiveness of a visual is completely
dependent on *the question you are trying to answer*. A visual that is
totally ineffective for one question may be perfect for answering a
different question.

In this challenge, you will practice *iterating* on data visualization,
and will anchor the *assessment* of your visuals using two different
questions.

*Note*: Please complete your initial visual design **alone**. Work on
both of your graphs alone, and save a version to your repo *before*
coming together with your team. This way you can all bring a diversity
of ideas to the table!

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|------------------|-----------------------------|-------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight**
before the day of the class discussion of the challenge. See the
[Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true)
for more information.

```{r setup}
library(tidyverse)
library(ggrepel)
```

*Background*: The data[1] we study in this challenge report the
[*minimum inhibitory
concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration)
(MIC) of three drugs for different bacteria. The smaller the MIC for a
given drug and bacteria pair, the more practical the drug is for
treating that particular bacteria. An MIC value of *at most* 0.1 is
considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin,
streptomycin, and neomycin---on 16 bacteria. Bacteria are categorized
into a genus based on a number of features, including their resistance
to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics %>% knitr::kable()
```

# Visualization

<!-- -------------------------------------------------- -->

### **q1** Prototype 5 visuals

To start, construct **5 qualitatively different visualizations of the
data** `df_antibiotics`. These **cannot** be simple variations on the
same graph; for instance, if two of your visuals could be made identical
by calling `coord_flip()`, then these are *not* qualitatively different.

For all five of the visuals, you must show information on *all 16
bacteria*. For the first two visuals, you must *show all variables*.

*Hint 1*: Try working quickly on this part; come up with a bunch of
ideas, and don't fixate on any one idea for too long. You will have a
chance to refine later in this challenge.

*Hint 2*: The data `df_antibiotics` are in a *wide* format; it may be
helpful to `pivot_longer()` the data to make certain visuals easier to
construct.

#### Visual 1 (All variables)

In this visual you must show *all three* effectiveness values for *all
16 bacteria*. You must also show whether or not each bacterium is Gram
positive or negative.

```{r q1.1}
df_antibiotics %>%
  pivot_longer(
    cols = c(penicillin, streptomycin, neomycin),
    names_to = "type",
    values_to = "value"
  ) %>%
  ggplot() +
  geom_jitter(aes(x = type, y = value, color = gram, shape = type)) +
  facet_wrap(~gram, scales = "free_y") +
  theme(legend.position = "none")
```

#### Visual 2 (All variables)

In this visual you must show *all three* effectiveness values for *all
16 bacteria*. You must also show whether or not each bacterium is Gram
positive or negative.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.2}
df_antibiotics %>%
  pivot_longer(
    cols = c(penicillin, streptomycin, neomycin),
    names_to = "type",
    values_to = "value"
  ) %>%
  ggplot() +
  geom_col(aes(x = type, y = value, color = type, fill = type)) +
  facet_wrap(~bacteria, scales = "free_y") +
  theme(axis.text.x=element_blank())
```

#### Visual 3 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.3}
df_antibiotics %>%
  ggplot() +
  geom_point(aes(x = streptomycin, y = neomycin, color = gram, size = penicillin))
```

#### Visual 4 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.4}
df_antibiotics %>%
  mutate(avg = rowSums(df_antibiotics[, c(2, 3, 4)], na.rm=TRUE) / 3) %>%
  ggplot() +
  geom_point(aes(x = bacteria, y = avg)) +
  coord_flip() +
  ggtitle("Average MICs across penicillin, streptomycin, & neomycin")
```

#### Visual 5 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.5}
df_antibiotics %>%
  pivot_longer(
    cols = c(penicillin, streptomycin, neomycin),
    names_to = "type",
    values_to = "value"
  ) %>%
  separate(bacteria, c("genus", "species"), " ") %>%
  ggplot() +
  geom_col(aes(x = type, y = value, color = gram, fill = gram)) +
  facet_wrap(~genus, scales = "free_y")
```

### **q2** Assess your visuals

There are **two questions** below; use your five visuals to help answer
both Guiding Questions. Note that you must also identify which of your
five visuals were most helpful in answering the questions.

*Hint 1*: It's possible that *none* of your visuals is effective in
answering the questions below. You may need to revise one or more of
your visuals to answer the questions below!

*Hint 2*: It's **highly unlikely** that the same visual is the most
effective at helping answer both guiding questions. **Use this as an
opportunity to think about why this is.**

#### Guiding Question 1

> How do the three antibiotics vary in their effectiveness against
> bacteria of different genera and Gram stain?

-   What is your response to the question above?

    -   Penicillin negative is extremely ineffective against 4 strains
        of bacteria (much more so than any other antibiotic and gram
        stain combination). This is indicated by the high MIC values -
        well over 750. In comparison to the rest of the data (most with
        MIC values under 100), this is massive.

    -   Penicillin positive, on the other hand, appears to be quite
        effective against all the bacteria it was tested on. Many of the
        data points appear to fall at or very close to 0. None of the
        points are visually extreme outliers like the four in the
        penicillin negative plot.

    -   Streptomycin positive has points that appear to all lie quite
        closely to 0, which supposedly indicates that streptomycin
        positive is effective against the bacteria it was tested on. I
        will make a note though that the visualization I am referencing
        has an oversized scale, up to probably 900, compared to the
        points I am looking at, which are probably less than 10. This
        makes it extremely difficult to see how much these points vary,
        and whether they are actually close to the 0.1 threshold or not.

    -   Streptomycin negative appears to have quite a large spread of
        values. One bacteria's MIC value is at 40, a couple are at 10,
        and the rest appear to be quite close to 0. Because there are a
        few points that are not visually close to 0, it can be said that
        streptomycin is not a consistently effective treatment against
        these bacteria with a negative stain.

    -   Neomycin positive, much like Streptomycin positive, has points
        that all appear to be close to 0. This implies that Neomycin
        positive is an effective treatment to all the bacteria. However,
        it is again important to point out that the y scale of the graph
        is much higher than whatever scale the Neomycin positive MIC
        values are operating on. This makes it hard to analyze
        differences of 10 or less.

    -   Neomycin negative is pretty effective for most bacteria, as many
        of the MIC values are close to 0. Only 2 bacteria, it seems,
        have high values close to 10 that indicate a lack of
        effectiveness.

-   Which of your visuals above (1 through 5) is **most effective** at
    helping to answer this question? Why?

    -   Visual 1 was most effective at helping me answer this question.
        It separated by both the three antibiotics and the gram stain,
        which allowed me to then easily see each combination's
        effectiveness across the bacteria it was tested on. However, I
        didn't add anything within the visual 1 plot to distinguish by
        genus, which is why I tweaked my visual 5 to directly assist
        with this.

#### Guiding Question 2

In 1974 *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*,
and in 1984 *Streptococcus fecalis* was renamed *Enterococcus fecalis*
[2].

> Why was *Diplococcus pneumoniae* was renamed *Streptococcus
> pneumoniae*?

*Observations*

-   What is your response to the question above?

    -   Perhaps because *Diplococcus pneumoniae* shares a similar
        pattern of MIC values across the three antibiotics as
        *Streptococcus hemolyticus* and *Streptococcus viridans*. In
        fact, *Diplococcus pneumoniae* not only has a similar pattern as
        *Streptococcus hemolyticus*, but they also have similar
        *values*. All three of these bacteria have quite high MIC values
        for streptomycin and neomycin, and a low MIC value for
        penicillin. It makes sense that scientists would want to group
        similarly-behaving bacteria by name, and therefore change
        *Diplococcus pneumoniae* to be *Streptococcus pneumoniae*.

-   Which of your visuals above (1 through 5) is **most effective** at
    helping to answer this question? Why?

    -   #2, because it allows me to see the patterns that each specific
        bacteria exhibits. Because I used a column plot, the blocks of
        color make it really easy to see which bacteria behave
        similarly. This allowed me to quickly notice that *Diplococcus
        pneumoniae* has a similar behavior to all three antibiotics as
        *Streptococcus hemolyticus* and *Streptococcus viridans* do.

# References

<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide
antibacterial range and rarely sensitizing. Scope. 1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." *American Scientist* (2009)
[link](https://www.americanscientist.org/article/thats-funny)
